Miniproject - Algebraic Geometry

### 2021-12-06 Algebraic geometry/submission.tex
proved the isomorphism

\begin{definition}
    Let $V$ be a vector space.
    Then define the $d$-th exterior power as
    \begin{equation*}
        \extpow^d(V) := V^{\otimes d} \ / \ \sum_{i = 1}^{d - 1} V^{\otimes (i - 1)} \otimes \bigl\{ v \otimes v' + v' \otimes v \ \bigm| \ v, v' \in V \bigr\} \otimes V^{\otimes (d - i - 1)}
    \end{equation*}
    Use the notation $v_1 \wedge ... \wedge v_d := [v_1 \otimes ... \otimes v_d] \in \extpow^k(V)$.
\end{definition}

@@

@@

    Note that for any nonzero decomposable vector
    \begin{equation*}
        u_1 \wedge u_2 \in \Bigl( \extpow^2 \vspan\{v_2, ..., v_t\} \Bigr) \setminus \{0\}
    \end{equation*}
    find
    \begin{equation*}
        u_1, u_2 \in \vspan\{v_2, ..., v_t\}
    \end{equation*}
    In particular, we know that
    \begin{equation*}
        v_1 \wedge \Bigl( \sum_{j > i} v_j \Bigr) \in \extpow^2 \vspan\{v_2, ..., v_t\}
    \end{equation*}
    and so $v_1 \in \vspan\{v_2, ..., v_t\}$ unless $\sum_{j > i} v_j = 0$.
    We assumed that the $v_i$ are linearly independent, so the former would give a contradiction.
    Hence $\sum_{j > i} v_j = 0$ and thus $t = 1$, i.e. $\omega = v_1$ is decomposable.
\end{proof}

\begin{lemma}[2a]
    \label{prop:linear_transform_extpow}
    Let $A = (a_{ij}) \in \GL_d(k)$ and $v_1, ..., v_d \in V$.
    Then
    \begin{equation*}
        \Bigl( \sum_j a_{1j}v_j \Bigl) \wedge ... \wedge \Bigl( \sum_{j} a_{dj} v_j \Bigr) = \det(A) (v_1 \wedge ... \wedge v_d)
    \end{equation*}
\end{lemma}
\begin{proof}
    By a direct computation using Lemma~\ref{prop:basic_properties_exterior_product}, we find
    \begin{align*}
        &\Bigl( \sum_j a_{ij} v_j \Bigr) \wedge ... \wedge \Bigl( \sum_j a_{dj} v_j \Bigr) = \sum_{j_1, ..., j_d} a_{1j_1} ... a_{dj_d} (v_{j_1} \wedge ... \wedge v_{j_d}) \\
        = &\sum_{\pi \in S_d} a_{1\pi(1)} ... a_{d\pi(d)} (v_{\pi(1)} \wedge ... \wedge v_{\pi(d)}) \\
        = &\sum_{\pi \in S_d} a_{1\pi(1)} ... a_{d\pi(d)} \sgn(\pi) (v_1 \wedge ... \wedge v_d) \\
        = &(v_1 \wedge ... \wedge v_d) \ \sum_{\pi \in S_d} \sgn(\pi) \prod_{j = 1}^d a_{j\pi(j)} = \det(A) (v_1 \wedge ... \wedge v_d)
    \end{align*}
    where the last equality holds due to the Leibniz determinant formula.
\end{proof}
We see that if there are two bases $v_1, ..., v_d$ and $u_1, ..., u_d$ of a $d$-dimensional vector space $U$, then there exists a basis change matrix $A = (a_{ij}) \in \GL_d(k)$ with
\begin{equation*}
    u_i = \sum_j a_{ij} v_j
\end{equation*}
So by the Lemma, it follows that
\begin{equation*}
    u_1 \wedge ... \wedge u_d = \det(A) (v_1 \wedge ... \wedge v_d)
\end{equation*}
As $v_1, ..., v_d$ resp. $u_1, ..., u_d$ are bases, they are linearly independent and in particular, we see that
\begin{equation*}
    v_1 \wedge ... \wedge v_d \neq 0 \quad \text{and} \quad u_1 \wedge ... \wedge u_d \neq 0
\end{equation*}
Hence they have well-defined images $[v_1 \wedge ... \wedge v_d]$ resp. $[u_1 \wedge ... \wedge u_d]$ in the projective space $\Proj(\extpow^d V)$.
By the above, find
\begin{equation*}
    [v_1 \wedge ... \wedge v_d] = [u_1 \wedge ... \wedge u_d]
\end{equation*}
This allows us to study the Grassmanian $\Gr(d, V)$ of a fixed vector space $V$.
\begin{definition}
    Define the map
    \begin{equation*}
        \phi: \Gr(d, V) \to \Proj(\extpow^d V), \quad \vspan\{v_1, ..., v_d\} \mapsto [v_1 \wedge ... \wedge v_d]
    \end{equation*}
    which is well-defined by Lemma~\ref{prop:linear_transform_extpow} as described above.
\end{definition}
\begin{lemma}[1a]
    We have
    \begin{equation*}
        \mathrm{im}\phi = D := \{ [v] \in \Proj(\extpow^d V) \ | \ \text{$v$ decomposable}\}
    \end{equation*}
\end{lemma}
\begin{proof}
    First of all, note that the set $D$ is well-defined, as $v$ is decomposable if and only if $\lambda v$ is decomposable, for all $\lambda \in k^*$.

@@

    Assume not, then there is a nonzero vector $a_1 \in k^d$ with $\sum a_{1i} v_i = 0$.
    Clearly, we can extend $a_1$ to a basis $a_1, ..., a_d$ of $k^d$, which gives a matrix $A = (a_{ij}) \in \GL_d(k)$.

    However by Lemma~\ref{prop:linear_transform_extpow} we now get
        0 =& 0 \wedge \Bigl( \sum_j a_{2j} v_j \Bigr) \wedge ... \wedge \Bigl( \sum_j a_{dj} v_j \Bigr) = \Bigl( \sum_j a_{1j} v_j \Bigr) \wedge ... \wedge \Bigl( \sum_j a_{dj} v_j \Bigr) \\
        =& \det(A) (v_1 \wedge ... \wedge v_d)
    and so $v = v_1 \wedge ... \wedge v_d = 0$ as $\det(A) \neq 0$.
    However, $v$ was a representative of a point in $\Proj(\extpow^d V)$, a contradiction.
    The claim follows.

@@

    where $v_1 \otimes ... \otimes v_d := \mathbb{1}_{(v_1, ..., v_d)}$.

@@

    It is clear by definition that $f$ is linear, so it is left to show that it is bijective. 
\end{proof}
\end{document}

>>>>

\section{Part I}

\begin{definition}
    Let $V$ be a vector space.
    Then define the $d$-th exterior power as
    \begin{equation*}
        \extpow^d(V) := V^{\otimes d} \ / \ \sum_{i = 1}^{d - 1} V^{\otimes (i - 1)} \otimes \vspan\bigl\{ v \otimes v' + v' \otimes v \ \bigm| \ v, v' \in V \bigr\} \otimes V^{\otimes (d - i - 1)}
    \end{equation*}
    Use the notation $v_1 \wedge ... \wedge v_d := [v_1 \otimes ... \otimes v_d] \in \extpow^k(V)$.
\end{definition}

@@

\begin{lemma}
    \label{prop:linear_transform_extpow}
    Let $A = (a_{ij}) \in \GL_d(k)$ and $v_1, ..., v_d \in V$.
    Then
    \begin{equation*}
        \Bigl( \sum_j a_{1j}v_j \Bigl) \wedge ... \wedge \Bigl( \sum_{j} a_{dj} v_j \Bigr) = \det(A) (v_1 \wedge ... \wedge v_d)
    \end{equation*}
\end{lemma}
\begin{proof}
    By a direct computation using Lemma~\ref{prop:basic_properties_exterior_product}, we find
    \begin{align*}
        &\Bigl( \sum_j a_{ij} v_j \Bigr) \wedge ... \wedge \Bigl( \sum_j a_{dj} v_j \Bigr) = \sum_{j_1, ..., j_d} a_{1j_1} ... a_{dj_d} (v_{j_1} \wedge ... \wedge v_{j_d}) \\
        = &\sum_{\pi \in S_d} a_{1\pi(1)} ... a_{d\pi(d)} (v_{\pi(1)} \wedge ... \wedge v_{\pi(d)}) \\
        = &\sum_{\pi \in S_d} a_{1\pi(1)} ... a_{d\pi(d)} \sgn(\pi) (v_1 \wedge ... \wedge v_d) \\
        = &(v_1 \wedge ... \wedge v_d) \ \sum_{\pi \in S_d} \sgn(\pi) \prod_{j = 1}^d a_{j\pi(j)} = \det(A) (v_1 \wedge ... \wedge v_d)
    \end{align*}
    where the last equality holds due to the Leibniz determinant formula.
\end{proof}
\begin{lemma}
    \label{prop:extpow_zero_iff_independent}
    For $v_1, ..., v_d \in V$ have
    \begin{equation*}
        v_1 \wedge ... \wedge v_d = 0 \ \Leftrightarrow \ v_1, ..., v_d \ \text{linearly dependent}
    \end{equation*}
\end{lemma}
\begin{proof}
    For the direction $\Rightarrow$, assume that $v_1, ..., v_d$ are not independent. 
    Then there is a nonzero vector $a_1 \in k^d$ with $\sum a_{1i} v_i = 0$.
    Clearly, we can extend $a_1$ to a basis $a_1, ..., a_d$ of $k^d$, which gives a matrix $A = (a_{ij}) \in \GL_d(k)$.

    However by Lemma~\ref{prop:linear_transform_extpow} we now get
    \begin{align*}
        0 =& 0 \wedge \Bigl( \sum_j a_{2j} v_j \Bigr) \wedge ... \wedge \Bigl( \sum_j a_{dj} v_j \Bigr) = \Bigl( \sum_j a_{1j} v_j \Bigr) \wedge ... \wedge \Bigl( \sum_j a_{dj} v_j \Bigr) \\
        =& \det(A) (v_1 \wedge ... \wedge v_d)
    \end{align*}
    and so $v = v_1 \wedge ... \wedge v_d = 0$ as $\det(A) \neq 0$.

    Direction $\Leftarrow$ TODO
\end{proof}
\begin{lemma}
    \label{prop:subspace_closed}
    Let $v \in V$ and $u \in \extpow^{d - 1} U$ for a linear subspace $U \leq V$. 
    If $u \neq 0$ then $v \in U$.
\end{lemma}
\begin{proof}
    TODO
\end{proof}

@@

    Note that for any nonzero decomposable vector
    \begin{equation*}
        u_1 \wedge u_2 \in \Bigl( \extpow^2 \vspan\{v_2, ..., v_t\} \Bigr) \setminus \{0\}
    \end{equation*}
    find
    \begin{equation*}
        u_1, u_2 \in \vspan\{v_2, ..., v_t\}
    \end{equation*}
    In particular, we know that
    \begin{equation*}
        v_1 \wedge \Bigl( \sum_{j > i} v_j \Bigr) \in \extpow^2 \vspan\{v_2, ..., v_t\}
    \end{equation*}
    and so $v_1 \in \vspan\{v_2, ..., v_t\}$ unless $\sum_{j > i} v_j = 0$ by Lemma~\ref{prop:subspace_closed}.
    We assumed that the $v_i$ are linearly independent, so the former would give a contradiction.
    Hence $\sum_{j > i} v_j = 0$ and thus $t = 1$, i.e. $\omega = v_1$ is decomposable.
\end{proof}

\section{Part II}

In this part, we want to consider the connection of external powers to the Grassmanian.
First of all, assume there are two bases $v_1, ..., v_d$ and $u_1, ..., u_d$ of a $d$-dimensional vector space $U$. 
Then there exists a basis change matrix $A = (a_{ij}) \in \GL_d(k)$ with
\begin{equation*}
    u_i = \sum_j a_{ij} v_j
\end{equation*}
So by Lemma~\ref{prop:linear_transform_extpow}, it follows that
\begin{equation*}
    u_1 \wedge ... \wedge u_d = \det(A) (v_1 \wedge ... \wedge v_d)
\end{equation*}
As $v_1, ..., v_d$ resp. $u_1, ..., u_d$ are bases, they are linearly independent and in particular, we see that
\begin{equation*}
    v_1 \wedge ... \wedge v_d \neq 0 \quad \text{and} \quad u_1 \wedge ... \wedge u_d \neq 0
\end{equation*}
by Lemma~\ref{prop:extpow_zero_iff_independent}.
Hence they have well-defined images $[v_1 \wedge ... \wedge v_d]$ resp. $[u_1 \wedge ... \wedge u_d]$ in the projective space $\Proj(\extpow^d V)$.
By the above, find
\begin{equation*}
    [v_1 \wedge ... \wedge v_d] = [u_1 \wedge ... \wedge u_d]
\end{equation*}
This allows us to study the Grassmanian $\Gr(d, V)$ of a fixed vector space $V$.
\begin{definition}
    Define the map
    \begin{equation*}
        \phi: \Gr(d, V) \to \Proj(\extpow^d V), \quad \vspan\{v_1, ..., v_d\} \mapsto [v_1 \wedge ... \wedge v_d]
    \end{equation*}
    which is well-defined by Lemma~\ref{prop:linear_transform_extpow} as described above.
\end{definition}
\begin{lemma}[1a]
    We have
    \begin{equation*}
        \mathrm{im}\phi = D := \{ [v] \in \Proj(\extpow^d V) \ | \ \text{$v$ decomposable}\}
    \end{equation*}
\end{lemma}
\begin{proof}
    First of all, note that the set $D$ is well-defined, as $v$ is decomposable if and only if $\lambda v$ is decomposable, for all $\lambda \in k^*$.

@@

    This follows directly from Lemma~\ref{prop:extpow_zero_iff_independent}.

@@

    where $v_1 \otimes ... \otimes v_d := \chi_{(v_1, ..., v_d)}$.

@@

    It is clear by definition that $f$ is linear, so it is left to show that it is bijective. 
    To show surjectivity, note that the $\pm e_I, I \in \{1, ..., n\}^{(d)}$ form a basis of $k^{\{1, ..., n\}^{(d)}}$.
    Clearly for $I = \{i_1, ..., i_d\}, J \in \{1, ..., n\}^{(d)}$ we have that
    \begin{equation*}
        f(e_{i_1} \wedge ... \wedge e_{i_d})_J = \det(A_J(e_{i_1}, ..., e_{i_j})) = \begin{cases}
            0 & \text{if $J \not\subseteq I$} \\
            \pm 1 & \text{if $J \subseteq I$}
        \end{cases}
    \end{equation*}
    so $f(e_{i_1} \wedge ... \wedge e_{i_d}) = e_I$ and we deduce that $\mathrm{im}f = \Proj^{\{1, ..., k\}^{(d)}}$.

    Finally, note that
    \begin{equation*}
        e_{i_1} \wedge ... \wedge e_{i_d}
    \end{equation*}
    for $i_1 < ... < i_d$ form a basis of $\extpow^d k^n$.
    Clearly, they span $\extpow^d k^n$, and the following argument shows that they are linearly independent.
    Assume
    \begin{equation*}
        \sum_{i_1 < ... < i_d} \lambda_{i_1, ..., i_d} (e_{i_1} \wedge ... \wedge e_{i_d}) = 0
    \end{equation*}
    Then
    \begin{equation*}
        0 = e_1 \wedge \Bigl( \sum_{1 < i_2 < ... < i_d} \lambda_{1, i_2, ..., i_d} (e_{i_2} \wedge .. \wedge e_{i_d}) \Bigr) + \sum_{1 < i_1 < ... < i_d} \lambda_{i_1, ..., i_d} (e_{i_1} \wedge ... \wedge e_{i_d} )
    \end{equation*}
    Clearly $e_1 \notin \vspan\{e_2, ..., e_n\}$ and so by Lemma~\ref{prop:subspace_closed} we see that
    \begin{equation*}
        \sum_{1 < i_2 < ... < i_d} \lambda_{1, i_2, ..., i_d} (e_{i_2} \wedge ... \wedge e_{i_d}) = 0
    \end{equation*}
    Repeating this argument inductively shows that $\lambda_{1, 2, ..., d} = 0$.
    As $k^n$ is symmetric w.r.t. permuting the $e_j$, we see that all $\lambda_{i_1, ..., i_d} = 0$ are zero.

    It follows that $\dim(\extpow^d k^n) = \dim(\Proj^{\{1, ..., n\}^{(d)}})$ and we find that $f$ is also injective.
\end{proof}
\end{document}
